The Ranked Nearest Neighbour Interchange space

### 2019-05-20 paper.tex
proof caterpillar convexity (?)

\begin{lemma}
    Let $T$ be the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.\\
    A caterpillar tree $T'$ has distance $d(T,T') = \Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$ to $T$ if, and only if, $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$
\end{lemma}

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T_1 = (\ldots ((1,2),3), \ldots, n)$, $T_2 = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$ on $n$ taxa it is $d(T_1, T_2) = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, T'$ be trees on $n+1$ taxa as described in the Lemma and a shortest path $p$ from $T$ to $T'$.
    Consider the trees $T|_n$ and $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$, respectively.
    Deleting taxon $n+1$ from all trees on $p$, and deleting all trees that appear more than once in a row, results in a path $p'$ from $T|_n$ to $T'|_n$.
    Applying the induction hypothesis onto $T|_n$ and $T'|_n$ shows that this path has length $\frac{(n-1)(n-2)}{2}$.
    We can now compare this new path $p'$ with $p$.
    The only steps that appear on $p$ but not on $p'$ are those involving the additional internal node.
    We can label this node by $n+1$ and define a bijection between the internal nodes in all trees on $p'$ and the corresponding nodes in the trees in $p'$.
    The additional internal node $n+1$ has rank $n$ in $T$ and ran $1$ in $T'$.
    Each $\rnni$ move can decrease the rank of a node by at most $1$.
    Therefore, there are at least $n-1$ $\rnni$ moves needed on $p$, which do not appear on $p'$, for moving the internal node $n+1$ down.
    Hence, it is $d(T,T') = |p| \geq n-1 + |p'| = n-1 + \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2} = \Delta(\rnniu)$.

>>>>

\begin{lemma}
    Let $T$ be the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.\\
    A caterpillar tree $T'$ has distance $d(T,T') = \Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$ to $T$ if, and only if, $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$
    \label{lemma:caterpillar_dist=diameter}
\end{lemma}

\todo{At the top of $T'$ it could also be $2,1$, or $t=1$}

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T_1 = (\ldots ((1,2),3), \ldots, n)$, $T_2 = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$ on $n$ taxa it is $d(T_1, T_2) = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, T'$ be trees on $n+1$ taxa as described in the Lemma and a shortest path $p$ from $T$ to $T'$.
    Consider the trees $T|_n$ and $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$, respectively.
    Deleting taxon $n+1$ from all trees on $p$, and deleting all trees that appear more than once in a row, results in a path $p'$ from $T|_n$ to $T'|_n$.
    Applying the induction hypothesis onto $T|_n$ and $T'|_n$ shows that this path has length $\frac{(n-1)(n-2)}{2}$.
    We can now compare this new path $p'$ with $p$.
    The only steps that appear on $p$ but not on $p'$ are those involving the additional internal node.
    We can label this node by $n+1$ and define a bijection between the internal nodes in all trees on $p'$ and the corresponding nodes in the trees in $p'$.
    The additional internal node $n+1$ has rank $n$ in $T$ and ran $1$ in $T'$.
    Each $\rnni$ move can decrease the rank of a node by at most $1$.
    Therefore, there are at least $n-1$ $\rnni$ moves needed on $p$, which do not appear on $p'$, for moving the internal node $n+1$ down.
    Hence, it is $d(T,T') = |p| \geq n-1 + |p'| = n-1 + \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2} = \Delta(\rnniu)$.
    Thus, there is a shortest path from $T$ to $T'$ that only consists of caterpillar trees.

\begin{theorem}
    The set of caterpillar trees is convex.
\end{theorem}

\begin{proof}
    We prove the Theorem by induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T_1$ and $T_2$.

    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T_1$, $T_2$ with distance $d(T_1,T_2) = \Delta(\rnniu)$ there is a caterpillar path from $T_1$ to $T_2$ between these trees.

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pait of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
    Let $T_1$ and $T_2$ be caterpillar trees with caterpillar distance $l$ \todo{define 'caterpillar distance'}.
    If $l \neq \Delta(\rnniu)$, there must be a pair of taxa $x,y$ that are neighboured in $T_2$ and have the same order in $T_1$ and $T_2$, which means that $rank(p_x) > rank(p_y)$ or $rank(p_x) < rank(p_y)$ holds in both trees $T_1$ and $T_2$.
    Then it follows that the tree $T_2'$ that equals $T_2$, but with exchanged taxa $x$ and $y$, has distance $l+1$ to $T_1$.
    So there must be a shortest path from $T_2'$ to $T_1$ that only consists of caterpillar trees.
    Since it does not matter in which order the pairs of taxa that need to exchange their position do exchange on a path between two caterpillar trees, $T_2$ is on a shortest caterpillar path from $T_2'$ to $T_1$.
    Therefore, it must be $d(T_1, T_2) = l$.
